Protocol to increase accuracy and fidelity of Pollen Meta-genomic Barcoding using Angiosperms353; Case study using Corbiculae loads from wild Bumble Bees

### 2023-05-29 script/manuscript_MEE.Rmd
writing in word

--- 
title: Plant Metagenomic Barcoding using Angiosperms353 of Corbiculae from wild Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$, Emily J. Woodworth$^1$$^,$$^2$, 
abstract:  |  
  1) DNA Barcoding has been successful for the rapid identification of species in complex ecological assemblages. However, metabarcoding in the plant kingdom has been limited due to a lack of universal gene regions that works across all taxa, limiting the applications of eDNA and metagenomics in ecology.  
  2) To circumvent these limitations, we propose using more holistic approach which combines a multiple gene approach that incorporates existing plant occurrence databases, species distribution models, and phenology analyses to generate a list of likely candidate species to increase accuracy. Such an effort can help guide plant sample collection and library construction, particularly for projects with limited resources or local taxonomic knowledge. In addition, we demonstrate that building a custom candidate plant list based on known occurrences, ecological requirements, and phenology can improve computing efficiency, speed, and accuracy of metabarcoding. We tested this approach to identify the plant species contributing to the corbiculae pollen loads of bumble bees. To validate the efficiency of this approach we compare results of the DNA barcoding to expert-led pollen identification and field observations of plant-pollinator interactions that generated these pollen loads.   
  3) We show that the Angiosperms 353 probes, developed for phylogenomics, and which are currently being used in the largest ever plant systematic endeavor, offer significant promise to metagenomic approaches around the globe. The DNA barcoding of bumble bee corbiculae pollen loads was most accurate when combined with knowledge of what plant species were flowering in the plant community when they were collected.   
  4) Supplementing DNA barcoding data with ecological context is most accurate and powerful.  
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  word_document: default
  toc: no
  pdf_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

```{r Required Libraries, message=F, warning=F, results='hide'}
library(tidyverse)
library(bookdown)
library(knitr)
library(kableExtra)
#devtools::install_github("kpmainali/CooccurrenceAffinity")
suppressWarnings(library(kableExtra))
pNumC <- function(...) prettyNum(...,big.mark = ',')

@@

GOES HERE

@@

### 2.5.1 Bee-Flower Observations and Pollen Load Collection

@@

content goes here

@@

## 2.3 | Pollen Morphological identification

### 2.3.1 | Pollen Reference Library

To qualitatively identify, and quantitatively note, the plant species present in corbiculae loads microscopy was used. A pollen reference library of fuchsin-jelly stained grains which may be present in corbiculae loads of slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@beattie1971technique, @brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (@ogilvie2022shifting), as well as the RMBL Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non-accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used clustering techniques to supplement our subjective opinions of which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (Appendix 6). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner.

### 2.3.2 | Preperation of Pollen Corbiculae Loads

  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Circa* 0.5mm^2^ of pollen was placed onto a ~4mm^2^ fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with clear nail polish to prevent oxidation; all samples are noted in Appendix 4.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained a unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects (Appendix 7 & 8).

### 2.4 | Molecular Barcoding

@@

### 2.4.2 | Temporal Analyses

For assignment of reads to ecologically probabilistic species subsequent to BLAST, flowering time was used as a filter. To estimate the duration of dates in which plant species were flowering Weibull estimates of several phenological parameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5 Level 4 Ecoregions, or conditionally 6 ecoregions if enough records were not found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, we used the SnowUS data set (@iler2021conceptual, @tran2019cloud) from 2000-2017 were analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in fall. Herbarium records after the 3^rd^ quantile for melt, and the 1^st^ quantile for snow cover of these metrics were removed. Species with > 10 records had their Weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered, we used the initiation and cessation dates, respectively, as effective start and ends of flowering. These estimates were compared to a long-term observational study of flowering phenology 1974-2012 (@caradonna2014shifts), and the floral abundance data from 2015, using Kendall's tau.

### 2.5.2 | Barcode references library

All lab work was carried out at The Daniel F. and Ada L. Rice Plant Conservation Science Center at the Chicago Botanic Garden, Glencoe, Illinois, U.S.A.

#### 2.5.2.1 | Sampling Species for Barcoding

Using five years (2015-2020) of observational data on *Bombus* queen interactions with flowering plants at these studies sites, we identified the plant taxa most frequently visited by queens across all years. In order to capture more variability inherit in the 353 loci we sequenced the 12 most visited taxa twice using samples collected from one site within the Gunnison Basin River Drainage and one individual collected from another more distal population. In addition we included a congener - or a species from a closely related genus to serve as an outgroup for all 12 taxa. We also sequenced another 15 taxa of plants commonly visited by *Bombus* workers, based on the abundances, and immediate access to plant tissue, in the aforementioned data set  (Appendix 3). Plant collections were identified typically using a combination, of dichotomous keys and primary literature as required (@flora1993flora, @hitchcock2018flora,  @ackerfield2015flora, @lesica2012manual,  @cronquist1977intermountain, @allred2012flora, @jepson2022online, @mohlenbrock2002vascular).

#### 2.5.2.2 | Plant Genomic DNA Extraction

Plant genomic DNA was isolated from ~ 1 cm^2^ of leaf tissue from silica-gel dried or herbarium material using a modified cetyltrimethylammonium (CTAB) protocol (@doylesCTAB) that included two chloroform washes. DNA was quantified using a Nanodrop 2000 (Thermo Fisher Scientific, Waltham, Massachusetts, USA) and Qubit fluorometer (Thermo Fisher Scientific).

#### 2.5.2.3 | Pollen Genomic DNA Extraction

To extract genomic DNA from pollen a CTAB based protocol was modified from Lahlamgiahi et al. and Guertler et al. (2014, 2014). A SDS extraction buffer (350µL , 100mM Tris-HCl, 50 mM EDTA, 50 mM NaCl, 10% SDS v/v., pH 7.5) was added to the sampled followed by vortexing to allow dissolution of corbiculae. Pollen grains were then macerated with Kontes Pellet Pestles, and the tip of these washed with 130 µL of the SDS extraction buffer, samples were then incubated for 1 hour at 30°C. This was followed by the addition of 10% CTAB solution (450ul, of 20 mM Tris-Cl pH. 8.0, 1.4 M NaCl, 10 mM EDTA pH 7.5, 10% CTAB, 5% PVP, ~85% Deionized water) and RNAse (10 uL of 10 mg/mL) and samples were incubated for 40 minutes at 37°C, on a heat block (Multi-Blok, Thermo Fisher Scientific, Waltham Massachusetts) set to 40°C. After 20 minutes incubation, Proteinase K (15 µL of 20mg/ml) and  DTT (12.5 µL of 1M in water) were added, and the samples were further incubated at 60°C for 1 hour. Samples were then incubated overnight at 40°C. 500 µL of Phenol-Chloroform-Isoamyl alcohol (25:24:1) were added, vortexed, and centrifuged at 10,000 rpm for 10 minutes and the aqueous phase was pipetted to a 1.5 ml centrifuge tube.

To precipitate the DNA, chilled Isopropyl alcohol & 3 mM Sodium acetate (5:1) equivalent to $\frac{2}{3}$ of the volume of sample were added, with 1 hour of chilling at -20°C, followed by 10 minutes of centrifuging at 13,000 rpm. The supernatant was pipetted to a new 1.5 ml centrifuge tube, and 70% EtOH (400 µL) were added before chilling at -20°C for 20 minutes followed by centrifugation at 13,000 rpm for 10 minutes. Both tubes were then washed with  75% EtOH (400 µL), inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded, then washed with 95% EtOH (400 µL), inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded. Pellets were dried at room temperature overnight before resuspension in nuclease free H~2~O.  Extractions were assessed using a Nanodrop 2000 (Thermo Fisher Scientific) and Qubit fluorometer (Thermo Fisher Scientific). DNA extracts were then cleaned using 2:1 v./v. Sera-Mag beads (Cytiva, Little Chalfont, UK) to solute ratio following the manufacturer's protocol, eluted in 0.5x TE, and the eluent allowed to reduce by half volume in ambient conditions. DNA was quantified using a Qubit fluorometer.

#### 2.5.2.4 | Fragmentation, Library Preparation & Target Enrichment

Sequence library preparation was performed using the NEBNext Ultra II FS-DNA Library Prep Kit for Illumina (New England BioLabs, Ipswich, Massachusetts, USA) using slightly modified manufacturers recommendation.  Fragmentation was performed at ½ volume of reagents and ¼ enzyme mix for 40 minutes at 37°C, with an input of 500 ng cleaned DNA. Adapter Ligation and PCR enrichment were performed with ½ volumes, while cleanup of products was performed using SPRI beads (Beckman Coulter, Indianapolis, Indiana, USA) and recommended volumes of 80% v./v. ethanol washes. The exception was the herbarium specimens which were not fragmented and only end repaired, with similar library preparation of all samples. Products were analysed on 4% agarose gels, and a Qubit fluorometer. 
Libraries were pooled and enriched with the Angiosperms 353 probe kit V.4 (Arbor Biosciences myBaits Target Sequence Capture Kit) by following the manufacturer's protocol and Brewer et al. 2019. Sequencing was performed using an Illumina mi-Seq with 150-bp end reads, (NUSeq Core, Chicago, Illinois).

###  2.6 | Computational Processes and Analyses. 
####  2.6.1 | Reference Library Data Processing

Sequences were processed using Trimmomatic, which removed sequence adapters, clipped the first 3 bp, discarding reads less than 36 bp, and removing reads if their average PHRED score dropped beneath 20 over a window of 5 bp (@bolger2014trimmomatic, @tange_2022_6377950). Contigs generated were mapped to a reference with HybPiper with using target files created by M353 (@johnson2016hybpiper, @mclay2021new).

#### 2.6.2 | Sequence Identification

A custom Kraken2 database was created by downloading representative species indicated as being present in the study area by the spatial analyses from the Sequence Read Archive (SRA) NCBI (@wood2019improved). These sequences were processed in the same manner as our novel sequences. The Kraken2 database was built using default parameters. Kraken2 was run on sequences using default parameters (Appendix 9).  Following Kraken2, Bracken was used to classify sequences to terminal taxa (@lu2017bracken). FInally all reads which could be classified by these databases were passed to a local BLAST database.

#### 2.6.3 | Identification of Sequence Matching Loci

A local NCBI database was built using the same processed novel and downloaded sequences as the previous database (@camacho2009blast).

@@

## 3.1 | Floral Observations

```{r Most used resources by species, fig.cap = 'Number of the ten most commonly visited plants which are also in the top ten most common sequences', out.width = "45%"}

@@

## 3.2 | Microscopic Pollen identification

@@

## 3.6 | Integrated Observational, Molecular, and Palynological Results

@@

expert %>% 
  group_by(Taxon) %>% 
  add_count(name = 'No.Corb') %>% 
  mutate(median_percent_corb = round( mean(Percent_Sample), 1), 
         Taxon = str_remove(Taxon, '^.* '), 
         Taxon = if_else(Taxon %in% c('Salix', 'Erigeron'), 'sp.', Taxon)) %>% 
  select(Genus, Species = Taxon, Percent_Sample, No.Corb, median_percent_corb) %>% 
  left_join(., fam, by = 'Genus') %>% 
  select(-Percent_Sample) %>% 
  relocate(Family, .before = Genus) %>% 
  filter(No.Corb > 4) %>% 
  distinct() %>% 
  arrange(Family, Genus)  %>% 
  knitr::kable(
    escape = F,
    align = c('l', 'l', 'l', 'c', 'c'),
    caption = 'Plant species detected in five or more corbiculae loads',
    col.names = c('Family', 'Genus', 'Species', 'No. Samples', 'Mean Prop.\\textsuperscript{a}')
  ) %>% 
  add_footnote(
    'The mean only calculated across the samples where the species was detected') %>% 
  collapse_rows()

@@

>>>>

--- 
title: Protocol to increase accuracy and fidelity of Pollen Meta-genomic Barcoding using Angiosperms353; Case study using Corbiculae loads from wild Bumble Bees
author:  |
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$, Jane E. Ogilive$^3$, 
    |  Sophie Taddeo$^1$$^,$$^2$$^,$$^4$, Emily J. Woodworth$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ The University of California at San Diego, 9500 Gilman Dr, La Jolla, California 92093, USA
abstract:  |  
  1) DNA Barcoding has been successful for the rapid identification of species in complex ecological assemblages. However, metabarcoding in the plant kingdom has been limited due to a lack of universal gene regions that works across all taxa, limiting the applications of eDNA and metagenomics in ecology.  
  2) To circumvent these limitations, we propose using more holistic approach which combines a multiple gene approach that incorporates existing plant occurrence databases, species distribution models, and phenology analyses to generate a list of likely candidate species to increase accuracy. Such an effort can help guide plant sample collection and library construction, particularly for projects with limited resources or local taxonomic knowledge. In addition, we demonstrate that building a custom candidate plant list based on known occurrences, ecological requirements, and phenology can improve computing efficiency, speed, and accuracy of metabarcoding. We tested this approach to identify the plant species contributing to the corbiculae pollen loads of bumble bees. To validate the efficiency of this approach we compare results of the DNA barcoding to expert-led pollen identification and field observations of plant-pollinator interactions that generated these pollen loads.   
  3) We show that the Angiosperms 353 probes, developed for phylogenomics, and which are currently being used in the largest ever plant systematic endeavor, offer significant promise to metagenomic approaches around the globe. The DNA barcoding of bumble bee corbiculae pollen loads was most accurate when combined with knowledge of what plant species were flowering in the plant community when they were collected.   
  4) Supplementing DNA barcoding data with ecological context is most accurate and powerful.  
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

```{r Required Libraries, message=F, warning=F, results='hide'}
library(tidyverse)
library(bookdown)
library(knitr)
library(kableExtra)
# devtools::install_github("kupietz/kableExtra") you may need this patch if error persists on render may. 27. 2023
#devtools::install_github("kpmainali/CooccurrenceAffinity")
suppressWarnings(library(kableExtra))
pNumC <- function(...) prettyNum(...,big.mark = ',')

@@

In order to acquire references genomes in a cost-effective manner, we suggest combining existing resources which are readily available at several databases, with *de novo* sequencing of species which are likely to be involved in the interactions of interest.

Although several online genomic databases do contain numerous taxa, these often contain only a single representative of each genus, and do not always account for geographic variability. For this reason, we are recommending using the reduced list of potential candidate plant species as a target list to collect leaf samples to generate a reference library. This will increase efficiency and accuracy of analyses down the line. We recognize that this might not be feasible for all study systems, or for extensively long candidate species list, however using current knowledge of the system and key players in an interaction will allow for more accurate hypothesis testing. For example, if interested in pollination, you can focus on known candidate species that represent flower morphologies associated with specific pollinator guilds (Fenster et al 2004). DNA extraction and processing of these should be done alongside the samples collected for this system.

@@

### Bee-Flower Observations and Pollen Load Collection

@@

#### Spatial Analyses

@@

@@

@@

```{r Most used resources by species, out.width = "45%"}

@@

@@

@@

expert %>% 
  group_by(Taxon) %>% 
  add_count(name = 'No.Corb') %>% 
  mutate(median_percent_corb = round( mean(Percent_Sample), 1), 
         Taxon = str_remove(Taxon, '^.* '), 
         Taxon = if_else(Taxon %in% c('Salix', 'Erigeron'), 'sp.', Taxon)) %>% 
  select(Genus, Species = Taxon, Percent_Sample, No.Corb, median_percent_corb) %>% 
  left_join(., fam, by = 'Genus') %>% 
  select(-Percent_Sample) %>% 
  relocate(Family, .before = Genus) %>% 
  filter(No.Corb > 4) %>% 
  distinct() %>% 
  arrange(Family, Genus)  %>% 
  knitr::kable(
    escape = F,
    align = c('l', 'l', 'l', 'c', 'c'),
    caption = 'Plant species detected in five or more corbiculae loads',
    col.names = c('Family', 'Genus', 'Species', 'No. Samples', 'Mean Prop.\\textsuperscript{a}')
  ) %>% 
  collapse_rows() %>% 
  add_footnote(
    'The mean only calculated across the samples where the species was detected')

@@

### Metabarcoding Pollen Identification

content should be here

### Validation

```{r Most used resources by species figure, fig.cap = 'Number of the ten most commonly visited plants which are also in the top ten most common sequences', out.width = "45%"}

## Microscopic Pollen identification
